Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-06 ms.tex
fixing the unfixable

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of a better alternative.

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Our justification for neglecting rotation will be discussed further in $\S\ref{sec:rotation}$.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the steady-state inflow/outflow solution (assuming one exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.  Also note that our solutions can be scaled to any value of the mass input parameter, $\eta$, since the mass and energy source terms scale linearly with $\rho$ or $\rho_{\star}$; however, the precise value of $\eta$ must be specified when cooling or thermal conduction are included, since these scale as gas density {\it squared}.  We check the accuracy of our numerical solutions by confirming that mass is conserved across the grid, in addition to the integral constraint on the energy (Bernoulli integral).

@@

If the SMBH dominates the gravitational potential at the stagnation radius, then $v_{w} \gtrsim \sigma$ and hence the stagnation radius resides well inside the sphere of influence ($r_{\rm s} < r_{\rm soi}$; eq.~[\ref{eq:rsoi}]).  In this limit, equation (\ref{eq:stag_analytic}) simplifies to
\begin{eqnarray}
  &\rs \underset{v_w \gtrsim \sigma}\approx \frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx \begin{cases}
    12
 \, \pc \,\, \Mbheight v_{500}^{-2}\, \pyear& \text{core} \\
    6
 \, \pc \,\, \Mbheight v_{500}^{-2} \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:stag_simple}
\end{eqnarray}
{\bf AG: missing a minus n here...}
where for the numerical estimate we adopt values of $n \approx 1$ and $n \approx 0.5$ for core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies, respectively, so chosen to match the results of our numerical calculations (Fig.~\ref{fig:stag}).  This expression is similar to that obtained by \citet{Volonteri+11} on more heuristic grounds (their eq.~6).  {\bf BDM: all analytic expressions have to be updated in core case to reflect this larger rs}

@@

Thermal stability ($\dot{q}_{\rm heat} > |\dot{q}_{\rm rad}|$; $t_{\rm
  cool} > t_{\rm ff}$) at the stagnation radius thus requires a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling3}
\end{align}
The requirement $v_{\rm w} \gtrsim \sigma$ results because the top two lines are derived using equation (\ref{eq:rs_simple}) for $r_s$.  However, when $v_{\rm w} \lesssim \sigma$ the stagnation radius moves to very large radii, or may not exist at all, resulting in much higher gas densities and inevitable cooling instability. 
%Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm CI} \sim 200-300$ km s$^{-1}$ is generally sufficient for stability at all radii (Fig.~\ref{fig:cooling}).

@@

\citet{EmsellemCappellari+:2007a} use two-dimensional kinematic data to measure the ratio of ordered to random motion in a sample of early type galaxies, which they quantify at each galactic radius $R$ by the parameter
  \begin{align}
    \lambda_R \equiv \frac{\langle r|V|\rangle}{\langle R\sqrt{V^2+\sigma^2}\rangle} \underset{R \ll r_{\rm soi}}\sim \frac{V_{\phi}}{\sigma},
  \end{align}
where $\sigma$ is the velocity dispersion and the brackets indicate a luminosity-weighted average.  The circularization radius of the accretion flow can be written in terms of $\lambda_R$ as
\be
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm soi}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\ee
where we have used the definition $r_{\rm soi} \equiv GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed that $r_{\rm s} \lesssim r_{\rm soi}$, as satisfied for thermally-stable solutions of interest.

\citet{EmsellemCappellari+:2007a} (their Fig.~2) find that $\lambda_R$ is generally $< 0.1$ on radial scales less that 10 per cent of the galaxy half-light radius and that $\lambda_R$ is a decreasing function of $R$.  From equation (\ref{eq:rcirc}) we conclude that $r_{\rm circ} \lesssim 0.01r_{s}$, a location well inside the inner sonic point of the flow (typically located at $\sim 0.1r_{\rm s}$; Fig.~\ref{fig:profiles}).  This suggests that angular momentum is only becomes important for thermally stable flows in a region that is causally disconnected from the rest of the flow.  We are thus justifying in neglecting the effects of rotation in calculating the mass accretion rate.

@@

Our numerical results, summarized in Table \ref{table:models} and Figure XXX, allow us to study a range of CNM properties and to assess the validity of the analytic estimates from the previous section.  Cases for which $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|< 1$ at the stagnation radius are marked as `thermally unstable', as are cases for which the stagnation radius diverges and hence could not be resolved, resulting in their calculations being terminated.

>>>>

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rsoi} \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rsoi}$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of a better alternative.

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the steady-state inflow/outflow solution (assuming one exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.  Also note that our solutions can be scaled to any value of the mass input parameter, $\eta$, since the mass and energy source terms scale linearly with $\rho$ or $\rho_{\star}$; however, the precise value of $\eta$ must be specified when cooling or thermal conduction are included, since these scale as gas density to a different power.  We check the accuracy of our numerical solutions by confirming that mass is conserved across the grid, in addition to the integral constraint on the energy (Bernoulli integral).

@@

If the SMBH dominates the gravitational potential at the stagnation radius, then $v_{w} \gtrsim \sigma$ and hence the stagnation radius resides well inside the sphere of influence ($r_{\rm s} < r_{\rm soi}$; eq.~[\ref{eq:rsoi}]).  In this limit, equation (\ref{eq:stag_analytic}) simplifies to
\begin{eqnarray}
  &\rs \underset{v_w \gtrsim \sigma}\approx \frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx \begin{cases}
    12
 \, \pc \,\, \Mbheight v_{500}^{-2}\, \pyear& \text{core} \\
    6
 \, \pc \,\, \Mbheight v_{500}^{-2} \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:stag_simple}
\end{eqnarray}
{\bf AG: missing a minus n here...}
where for the numerical estimate we adopt values of $n \approx 1$ and $n \approx 0.5$ for core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies, respectively, so chosen to match the results of our numerical calculations (Fig.~\ref{fig:stag}).  This expression is similar to that obtained by \citet{Volonteri+11} on more heuristic grounds (their eq.~6).

@@

Thermal stability ($\dot{q}_{\rm heat} > |\dot{q}_{\rm rad}|$; $t_{\rm
  cool} > t_{\rm ff}$) near the stagnation radius thus requires a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling3}
\end{align}
The requirement $v_{\rm w} \gtrsim \sigma$ results because the top two lines are derived using equation (\ref{eq:rs_simple}) for $r_s$.  However, when $v_{\rm w} \lesssim \sigma$ the stagnation radius moves to very large radii, or may not exist at all (see also Fig.~\ref{fig:stag}), resulting in much higher gas densities and inevitable cooling instability. 
%Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm CI} \sim 200-300$ km s$^{-1}$ is generally sufficient for stability at all radii (Fig.~\ref{fig:cooling}).

@@

\citet{EmsellemCappellari+:2007a} use two-dimensional kinematic data to measure the ratio of ordered to random motion in a sample of early type galaxies, which they quantify at each galactic radius $R$ by the parameter
  \begin{align}
    \lambda_R \equiv \frac{\langle r|V|\rangle}{\langle R\sqrt{V^2+\sigma^2}\rangle} \underset{R \ll r_{\rm soi}}\sim \frac{V_{\phi}}{\sigma},
  \end{align}
where $\sigma$ is the velocity dispersion and the brackets indicate a luminosity-weighted average.  The circularization radius of the accretion flow can be written in terms of $\lambda_R$ as
\be
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm soi}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\ee
where we have used the definition $r_{\rm soi} \equiv GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed that $r_{\rm s} \lesssim r_{\rm soi}$, a condition which is satisfied for the thermally-stable solutions of interest.

\citet{EmsellemCappellari+:2007a} (their Fig.~2) find that $\lambda_R$ is generally $< 0.1$ on radial scales $<$ 10 per cent of the galaxy half-light radius and that $\lambda_R$ is with $R$ interior to this point.  From equation (\ref{eq:rcirc}) we thus conclude that $r_{\rm circ} \lesssim 0.01r_{s}$, i.e.~that gas will circularize well inside the inner sonic point of the flow (typically located at $\sim 0.1r_{\rm s}$; Fig.~\ref{fig:profiles}).  Angular momentum should only become dynamically important for thermally stable flows in a region that is causally disconnected from the outer part of the flow where the mass accretion rate is determined.

@@

Our numerical results, summarized in Table \ref{table:models}, allow us to study a range of CNM properties and to assess the validity of the analytic estimates from the previous section.  Cases for which $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|< 1$ at the stagnation radius are marked as `thermally unstable', as are cases for which the stagnation radius diverges and hence could not be resolved, resulting in their calculations being terminated.
